---
title: "Calibrate Smoking Status"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calibrate Smoking Status}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This document outlines the steps taken to calibrate the model to align with U.S.-based validation targets for smoking behavior using a 25 year time horizon.

Validation Targes:
Smoking Rates: Summary Health Statistics, National Health Interview Survey (NHIS), 2018

### Overview

This section outlines the steps taken to calibrate smoking behavior and COPD prevalence in the model to align with U.S.-based validation targets.

Smoking Behavior Validation Targets (NHIS 2018):

Current smokers: 13.8%
Former smokers: 20.9%
Never smokers: 65.3%

### Step 1: Load libraries and setup

Here, we load the necessary libraries. We also set the default simulation settings and specify the time horizon for the simulation (25 years).

```{r, eval = TRUE, echo = TRUE, message=FALSE}
library(epicUS)
library(tidyverse)
library(ggplot2)
library(scales)
library(dplyr)
library(knitr)

# Load EPIC general settings
settings <- get_default_settings()
settings$record_mode <- 0
settings$n_base_agents <- 1e6
init_session(settings = settings)

input <- get_input()
time_horizon <- 26
input$values$global_parameters$time_horizon <- time_horizon

```

### Step 2: Modify intercept value to calibrate proportion of non-smokers and former smokeres in the model

EPIC's existing calibration function (from EPIC Canada) accurately matched the overall smoking rates from the 2018 Summary Health Statistics: National Health Interview Survey (NHIS), however, it underestimated the proportions of never and former smokers. To address this, we modified the intercept in the logistic regression equation that determines the probability of being a never smoker. Increasing the intercept to 4.85 improved alignment with the target distribution by increasing the proportion of former smokers.


```{r, eval = TRUE, echo = TRUE}
input$values$smoking$logit_p_never_smoker_con_not_current_0_betas<-t(as.matrix(c(intercept = 4.85, sex = 0, age = -0.06, age2 = 0,
                                                                                 sex_age = 0,sex_age2 = 0, year = -0.02)))
```

### Step 4: Run EPIC

```{r, eval = TRUE, echo = TRUE}

run(input = input$values)
output <- Cget_output_ex()
terminate_session()

```

### Step 5: Create data tables 

```{r, eval = TRUE, echo = TRUE}

#Calculate smoking proportions
smokingstatus_overall <- output$n_smoking_status_by_ctime
row_sums <- rowSums(smokingstatus_overall)
smokingstatus_proportions <- smokingstatus_overall / row_sums
smokingstatus_proportions <- as.data.frame(smokingstatus_proportions)

# Rename columns for readability
colnames(smokingstatus_proportions) <- c("Never Smoker", "Current Smoker", "Former Smoker")

# Add Year column
smokingstatus_proportions$Year <- 2015:2040

# Display summary of smoking status 
kable(
  head(smokingstatus_proportions[, c("Year", "Never Smoker", "Current Smoker", "Former Smoker")]),
  caption = "Proportion of Smoking Status Over Time",
  digits = 3
)
```

### Step 6: Visualize data

```{r,smokingstatus_reshaped, fig.width = 8, fig.height = 6, echo = TRUE}

# Reshape data for plotting
smokingstatus_reshaped<- pivot_longer(
  smokingstatus_proportions,
  cols = c("Never Smoker", "Current Smoker", "Former Smoker"),
  names_to = "Status",
  values_to = "Proportion"
)

# Define colors
poster_colors <- c(
  "Never Smoker" = "#003f5c",      # deep navy
  "Current Smoker" = "#ffa600",    # coral orange
  "Former Smoker" = "#66c2ff"      # light blue
)

# Plot smoking status trends
ggplot(smokingstatus_reshaped, aes(x = Year, y = Proportion, color = Status)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 3) +
  scale_color_manual(values = poster_colors) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 1)) +
  scale_x_continuous(breaks = seq(2015, 2040, by = 5)) +
  labs(
    title = "Smoking Status Trends Over Time",
    subtitle = "Smoking status from 2015–2040",
    x = "Year",
    y = "Proportion of Population",
    color = "Smoking Status"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 18, hjust = 0.5),
    plot.subtitle = element_text(size = 14, hjust = 0.5),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "black"),
    axis.line = element_line(color = "black", linewidth = 0.8),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 12, margin = margin(b = 4)),
    legend.key.height = unit(1.2, "lines")
  )

```

### Step 7: Observe trend in pack years per person

```{r, eval = TRUE, echo = TRUE}

# Calculate total pack years by year
pack_years_total<- as.data.frame(output$sum_pack_years_by_ctime_sex)
pack_years_total <- rowSums(pack_years_total)
pack_years_total <- data.frame(
  Year = 2015:2040,
  packyears = pack_years_total)

# Calculate total number of current and former smokers by year

smoking_history <- as.data.frame(output$n_smoking_status_by_ctime)
smoking_history <- rowSums(smoking_history [, 2:3])
smoking_history <- data.frame(
    Year = 2015:2040,
    n_smokers = smoking_history)

# Merge datasets by year
pack_years_total_and_smoking_history <- merge(pack_years_total, smoking_history, by = "Year")

# Calculate average pack-years 
pack_year_per_person <- data.frame(
  Year = pack_years_total_and_smoking_history$Year,
  AvgPackYearsPerSmoker = pack_years_total_and_smoking_history$packyears / pack_years_total_and_smoking_history$n_smokers
)

#
```

### Step 8: Visualize trend in pack years per person

```{r,pack_years_per_person, fig.width = 8, fig.height = 6, echo = TRUE}

ggplot(pack_year_per_person, aes(x = Year, y = AvgPackYearsPerSmoker)) +
  geom_line(linewidth = 1.5, color = "#003f5c") +           # Deep navy
  geom_point(size = 3, color = "#66c2ff", stroke = 0.8) +   # Light blue
  scale_y_continuous(
    labels = scales::comma_format(accuracy = 1),
    limits = c(0, NA)
  ) +
  scale_x_continuous(breaks = seq(2015, 2040, by = 5)) +
  labs(
    title = "Average Pack-Years per Smoker Over Time",
    subtitle = "Estimated smoking exposure per smoker from 2015–2040",
    x = "Year",
    y = "Average Pack-Years"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 18, hjust = 0.5, margin = margin(b = 8)),
    plot.subtitle = element_text(size = 14, hjust = 0.5),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "black"),
    axis.line = element_line(color = "black", linewidth = 0.8),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```

### 
