---
title: "Costs_and_resource_utilization"
output: html_document
---

### Overview

This document outlines the steps taken to calibrate the model to modify costs to the US setting

### Step 1: Calculating yearly costs of COPD medication

We utilized prescription fill data outlined in Mannino et al 2022 (DOI: 10.1016/j.rmed.2022.106807) for all COPD drug data except for SABA for which we used Bengtson et al 2018 (DOI:10.1177/1753466618772750). The studies estimated the following dispenses per patient per year

ICS/LAMA/LABA: 6.1 
ICS containing regimens: 5.3
LAMA containing regimens: 5.2
LABA containing regimens: 5.5
SABA: 1.0

To calculate the costs we used the average drug costs for each category of drugs

ICS/LAMA/LABA:
ICS:
LAMA LABA:

As this study utilizes real world data adherence factors have originally been considered thus we will not need to account or multiply against a adherence ratio.

```{r, eval = TRUE, echo = TRUE}

#COPD medication cost
input$values$medication$medication_costs <- c(None=0,SABA=133.8*input$values$medication$medication_adherence, LABA=0, SABA_LABA=0, LAMA=3050.4*input$values$medication$medication_adherence, LAMA_SABA=0, LAMA_LABA=5150.4*input$values$medication$medication_adherence, LAMA_LAMA_SABA=0, ICS=0, ICS_SABA=0, ICS_LABA=0, ICS_LABA_SABA=0, ICS_LAMA=0, ICS_LAMA_SABA=0, ICS_LAMA_LABA=7913.6*input$values$medication$medication_adherence, ICS_LAMA_LABA_SABA=0)

```

### Step 2: Estimating COPD-Related Background Costs by GOLD Stage

COPD-related background costs were estimated using data from Wallace et al 2019 (DOI: 10.18553/jmcp.2019.25.2.205) Table 3, specifically the row labeled "COPD-related costs, all patients." Background costs were calculated by subtracting the costs of **inpatient care**, **emergency room (ER) visits**, and **pharmacy** from the **total COPD-related medical costs**. 

**Formula:**

\[
\text{Background Cost} = \text{Total COPD Medical Cost} - (\text{Inpatient} + \text{ER} + \text{Pharmacy})
\]

**Results (2016 Costs USD):**

| GOLD Stage    | Total COPD Cost | Inpatient | ER Visits | Pharmacy | **Background Cost** |
|---------------|----------------:|----------:|----------:|---------:|--------------------:|
| GOLD I        | \$5,945          | \$3,853   | \$186     | \$592    | **\$1,314**         |
| GOLD II       | \$6,978          | \$4,449   | \$144     | \$1,101  | **\$1,284**         |
| GOLD III      | \$10,751         | \$6,277   | \$193     | \$2,000  | **\$2,281**         |
| GOLD IV       | \$18,070         | \$12,139  | \$534     | \$2,479  | **\$2,918**         |

```{r, eval = TRUE, echo = TRUE}

#Background COPD cost by GOLD stage
input$values$cost$bg_cost_by_stage=t(as.matrix(c(N=0, I=1738.28*1.0528, II=1698.59*1.0528, III=3017.52*1.0528, IV=3860.2*1.0528)))

```

### Step 3: Estimating exacerbation costs by severity

The exacerbation module assigns costs to events based on severity: **moderate**, **severe**, and **very severe**. These estimates were derived from U.S.-based healthcare cost studies and reflect per-event medical costs.

**Mild Exacerbations**  
Defined as exacerbations requiring emergency department visit (Dalal et al. 2011; DOI:10.1016/j.rmed.2010.09.003) 
**Cost:** $606 (2008 Costs USD)

**Moderate Exacerbations**  
Defined as exacerbations without hospitalization (Bogart et al. 2020; DOI: 10.37765/ajmc.2020.43157)  
**Cost:** $2107 (2017 Costs USD)

**Severe Exacerbations**  
Defined as exacerbations requiring inpatient hospitalization (Bogart et al. 2020; DOI: 10.37765/ajmc.2020.43157)  
**Cost:** $22,729 (2017 Costs USD)

**Very Severe Exacerbations**  
Defined as exacerbations requiring ICU admission and intubation (Dalal et al. 2011; DOI:10.1016/j.rmed.2010.09.003)  
**Estimated cost:** $44,909 (2008 Costs USD)

```{r, eval = TRUE, echo = TRUE}

#Background COPD cost by GOLD stage
input$values$cost$exac_dcost=t(as.matrix(c(mild=59.71*1.0528,moderate=942.23*1.0528,severe=10546.6*1.0528, verysevere=30228.64*1.0528)))

```

### Step 4: Estimating costs for smoking cessation 

According to the MMWR study (DOI: 10.15585/mmwr.mm7329a1) the most commonly used pharmacotherapies included the following and was used in people that wanted to quit (2025 Costs USD)

Nicotine patch = 19.6%
Nicotine gum/lozenge = 18.4%
Nicotine spray/inhaler = 1.0%
Varenicline = 9.6%
Bupropion = 6.4%

Reweighted to assume 100%

Nicotine patch	(19.6 / 55.0) × 100 ≈ 35.6%
Nicotine gum/lozenge	(18.4 / 55.0) × 100 ≈ 33.5%
Nicotine spray/inhaler	(1.0 / 55.0) × 100 ≈ 1.8%
Varenicline	(9.6 / 55.0) × 100 ≈ 17.5%
Bupropion	(6.4 / 55.0) × 100 ≈ 11.6%

The cost of the smoking cessation therapies were determined as follows, cost data was obtained from GoodRx.

Nicotine patch: $71 
Nicotine gum/lozenge: $35
Nicotine Spray: $550
Varenicline: $402
Bupropion: $25

Using these weights we then derive an average cost of smoking cessation for 3 months of $120.15

```{r, eval = TRUE, echo = TRUE}

input$cost$cost_smoking_cessation <- 120.15

```

### Step 4: Estimating costs for GP visit and diagnostic spirometry 

**GP visits**  
Medicare pays providers approximately $89 and $125 for a 99214 visit in a typical outpatient setting (Source: https://college.acaai.org/wp-content/uploads/2024/07/2025-Proposed-RVUs-and-Reimbursement-for-Allergy.pdf). A midpoint of $107 was used (2025 cost)

**Spirometry**  
CMS CPT code 94060 was used which equates to $41 (2025 Costs USD)

```{r, eval = TRUE, echo = TRUE}

#Cost of diagnostic spirometry
input$cost$cost_outpatient_diagnosis <- 41

#Cost of GP visits
input$cost$cost_gp_visit <- 107

```


