---
title: "Reproducing Background Mortality"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reproducing Background Mortality}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this document, we aim to explore how the model behaves when COPD-related mortality is turned off. Specifically, our goal is to observe probability of death values that closely align with the background mortality rates from official Canadian sources. This is the expected behavior, as disabling exacerbation-related deaths causes the model to no longer predict mortality from exacerbations, which should result in values matching the background mortality rates.

To achieve this, we followed the series of steps below:

### Step 1: Shutting off exacerbation related mortality.

First, we set the intercept of the exacerbation equation to a high value, effectively disabling death from exacerbations in the model. This adjustment is made by changing the relevant variable as follows:


```{r, eval = TRUE, echo = TRUE}
library(epicR)
packageVersion("epicR")
input<-get_input()
input$values$exacerbation$logit_p_death_by_sex <- cbind(
  male = c(intercept = -13000, age = log(1.05), mild = 0, moderate = 0, severe = 7.4,
    very_severe = 8, n_hist_severe_exac = 0),
  female = c(intercept = -13000, age = log(1.05), mild = 0, moderate = 0, severe = 7.4,
    very_severe = 8, n_hist_severe_exac = 0)
)

#also setting exlicit mortality = 0 so there is no correction made
input$values$manual$explicit_mortality_by_age_sex <- cbind(
    male = c(rep(0, 111)),
    female = c(rep(0, 111)))


```

### Step 2: Setting longevitiy-related parameters to 0.

Longevity is another factor in the model that influences population. To ensure that no external factors impact the population, we set these parameters to 0. This adjustment can be made in the input.R file.

```{r, eval = TRUE, echo = TRUE}
input$values$agent$ln_h_bgd_betas <- t(as.matrix(c(intercept = 0, y = 0, y2 = 0, age = 0,
                                            b_mi = 0, n_mi = 0, b_stroke = 0,
                                            n_stroke = 0, hf = 0)))

```

### Step 3: Run the model for 1 year and retrieve events matrix

Initially we run the model for 1 year and get the events matrix. This matrix logs all the events
that individuals go through thoughout the model. We will use this matrix to calculate death 
probabilities produced by epicR

```{r, eval = TRUE, echo = TRUE}
library(epicR)
settings <- get_default_settings()
settings$record_mode <- 2
settings$n_base_agents <- 3.5e5

init_session(settings = settings)
# input <- get_input()

# set time horizon as 1 initially 
time_horizon <- 1
input$values$global_parameters$time_horizon <- time_horizon

run(input = input$values)
events <- as.data.frame(Cget_all_events_matrix())
terminate_session()

# checking to make sure event 7 (death by exacerbation) is not included 
# because we shut that off in the model
unique(events$event)
table(events$event)

```
### Step 4: Calculating probability of death from events matrix

```{r, eval = TRUE, echo = TRUE, message=FALSE}
library(dplyr)
library(tidyr)

# we will group by age so we convert ages into whole numbers.
events <- events %>%
  mutate(age_and_local = floor(local_time + age_at_creation))

# Filter events to identify individuals who have experienced event 14,
# while also creating a flag for whether they ever experienced event 13 (death)
events_filtered<- events %>%
  mutate(death= ifelse(event==13,1,0)) %>%
  group_by(id) %>%
  mutate(ever_death = sum(death)) %>%
  filter(event==14) %>%
  ungroup()


# calculationg probability of death
death_prob<- events_filtered %>%
  group_by(age_and_local, female) %>%
  summarise(
    total_count = n(),
    death_count = sum(ever_death==1),
    death_probability = death_count / total_count
  )

```

### Inspecting the results

We should not consider 40 year-olds beause (Amin said ... )

```{r, eval = TRUE, echo = TRUE}
print(head(death_prob, 15))
print(tail(death_prob,15))
```

While these values are not perfectly aligned with our validation target, the variation is negligible.

Next, we want to ensure a consistent directional effect. We expect that increasing time_horizon
to 5 years will bring the results closer to our target.

```{r, eval = TRUE, echo = TRUE}

init_session(settings = settings)

# set time horizon to 5
time_horizon <- 6
input$values$global_parameters$time_horizon <- time_horizon

run(input = input$values)
events5 <- as.data.frame(Cget_all_events_matrix())
terminate_session()

table(events$event)

events5 <- events5 %>%
  mutate(age_and_local = floor(local_time + age_at_creation))


events5 <- events5 %>%
  mutate(local_time_adj = ceiling(events5$local_time))

# withing that year have they ever died?
events5_filtered<- events5 %>%
  mutate(death= ifelse(event==13,1,0)) %>%
  group_by(id,local_time_adj) %>%
  mutate(ever_death = sum(death)) %>%
  filter(event==14) %>%
  ungroup()


death_prob5<- events5_filtered %>%
  group_by(age_and_local, female, local_time_adj) %>%
  summarise(
    total_count = n(),
    death_count = sum(ever_death==1),
    death_probability = death_count / total_count
  )

```

Now, when we check the results. We see that the results are even further away:


```{r, eval = TRUE, echo = TRUE}
print(head(death_prob5, 15))
print(tail(death_prob5,15))
```

### Visualizing results

To better understand the differences, we visualize the results and compare them
against the target values.

```{r, eval = TRUE, echo = TRUE}
death_prob_clean <- death_prob %>%
   ungroup() %>%
   select(age_and_local, female, death_probability) %>%
   pivot_wider(names_from = female, values_from = death_probability, names_prefix = "sex_")

colnames(death_prob_clean) <- c("Age", "Male", "Female")

```

```{r, eval = TRUE, echo = TRUE}


Canadalifetables_num <- input$values$agent$p_bgd_by_sex


Canadalifetables_df <- data.frame(
  Age = 1:nrow(Canadalifetables_num),  # Start age from 1
  Male = Canadalifetables_num[, 1],
  Female = Canadalifetables_num[, 2]
)

common_ages <- intersect(death_prob_clean$Age, Canadalifetables_df$Age)

# filter both so only include the rows with matching Age
death_prob_filtered <- death_prob_clean[death_prob_clean$Age %in% common_ages, ]
Canadalifetables_filtered <- Canadalifetables_df[Canadalifetables_df$Age %in% common_ages, ]

```


```{r, fig.width=8, fig.height=5, eval = TRUE, echo = TRUE}
library(ggplot2)
library(dplyr)

combined_data_long <- bind_rows(
  death_prob_filtered %>% mutate(Source = "epicR"),
  Canadalifetables_filtered %>% mutate(Source = "Canada Life Tables")
) %>%
  pivot_longer(cols = c("Male", "Female"), names_to = "Sex", values_to = "Death_Probability") %>%
  filter(Age > 40)  

ggplot(combined_data_long, aes(x = Age, y = Death_Probability, fill = Source)) +
  geom_col(position = "dodge", width = 1) +  
  facet_wrap(~Sex) +
  labs(
    title = "Comparison of epicR Death Probability vs. Canada Life Tables (time_horizon =1)",
    x = "Age",
    y = "Death Probability",
    fill = "Source:"
  ) +
  theme_minimal()+
  theme(
    legend.position = "top",
    legend.justification = "center",
    plot.title = element_text(hjust = 0.5, margin = margin(b = 10))
  )


```

```{r, eval = TRUE, echo = FALSE}
# Reshape the death probability data to wide format for each year and sex
death_prob5_clean <- death_prob5 %>%
   ungroup() %>%
   select(age_and_local, female, death_probability, local_time_adj) %>%
   pivot_wider(names_from = female, values_from = death_probability, names_prefix = "sex_")

colnames(death_prob5_clean) <- c("Age", "Year", "Male", "Female")

# Filter the dataset based on age matching with the Canada life tables
common_ages <- intersect(death_prob5_clean$Age, Canadalifetables_df$Age)

# Filter both datasets to include only the common ages
death_prob5_filtered <- death_prob5_clean[death_prob5_clean$Age %in% common_ages, ]
Canadalifetables_filtered <- Canadalifetables_df[Canadalifetables_df$Age %in% common_ages, ]



```

When `time_horizon = 6`, we get the following plot.


```{r, fig.width=8, fig.height=5, eval = TRUE, echo = TRUE, warning=FALSE}

library(ggplot2)
library(tidyr)
library(dplyr)

combined_data_long5 <- bind_rows(
  death_prob5_filtered %>% mutate(Source = "epicR"),
  Canadalifetables_filtered %>% mutate(Source = "Canada Life Tables")
) %>%
  pivot_longer(cols = c("Male", "Female"), names_to = "Sex", values_to = "Death_Probability") %>%
  filter(Age > 40)  


final_EPIC_death5<- filter(combined_data_long5, Source == "epicR")
final_Canada_death<- filter(combined_data_long5, Source == "Canada Life Tables")

# Add Death_Probability to Canadalifetables_filtered
Canadalifetables_filtered_long <- Canadalifetables_filtered %>%
  gather(key = "Sex", value = "Death_Probability", Male, Female)

# Loop through each unique year in the combined_data_long dataset
for (year in unique(combined_data_long5$Year)) {
  
  # Filter the data for the current year
  year_data <- combined_data_long5 %>% filter(Year == year)
  
  # Create the plot for the current year
  p <- ggplot(combined_data_long, aes(x = Age, y = Death_Probability, fill = Source)) +
  geom_col(position = "dodge", width = 1) +  
  facet_wrap(~Sex) +
  labs(
    title = "Comparison of epicR Death Probability vs. Canada Life Tables (time_horizon = 6)",
    x = "Age",
    y = "Death Probability",
    fill = "Source:"
  ) +
  theme_minimal()+
  theme(
    legend.position = "top",
    legend.justification = "center",
    plot.title = element_text(hjust = 0.5, margin = margin(b = 10))
  )
  
  # Print the plot for the current year
  # print(p)
}

print(p)





```
When we break it down and analyze per year, it seems to be in line with what we expect.


### Eliminating smoking mortality

We have identified a potential source of the observed mortality discrepancies. Specifically, the mortality ratio associated with both former and current smokers appears to be influencing the results.

To investigate this, we set the following smoking-related mortality factors to 1 (indicating no excess mortality risk for former and current smokers):

After making this adjustment, the results align more closely with the life tables.

- This suggests that either:
  - The model may be generating an excessive number of current and former smokers, leading to higher mortality than expected.
  - We will need to update the mortality ratio estimates depending on whether an individual currently or formerly smokes.


```{r, eval=TRUE, echo=TRUE}


input$values$smoking$mortality_factor_former<- c(age40to49=1,age50to59=1,
                                                 age60to69=1,age70to79=1,
                                                 age80p=1)
input$values$smoking$mortality_factor_current<- c(age40to49=1,age50to59=1,
                                                  age60to69=1,age70to79=1,
                                                  age80p=1)
```

```{r, eval=TRUE, echo =FALSE, fig.width=8, fig.height=5}

init_session(settings = settings)
time_horizon <- 1
input$values$global_parameters$time_horizon <- time_horizon

run(input = input$values)
events_smoking <- as.data.frame(Cget_all_events_matrix())
terminate_session()

# we will group by age so we convert ages into whole numbers.
events_smoking <- events_smoking %>%
  mutate(age_and_local = floor(local_time + age_at_creation))

# Filter events to identify individuals who have experienced event 14,
# while also creating a flag for whether they ever experienced event 13 (death)
events_smoking_filtered<- events_smoking %>%
  mutate(death= ifelse(event==13,1,0)) %>%
  group_by(id) %>%
  mutate(ever_death = sum(death)) %>%
  filter(event==14) %>%
  ungroup()


# calculationg probability of death
death_prob_smoking<- events_smoking_filtered %>%
  group_by(age_and_local, female) %>%
  summarise(
    total_count = n(),
    death_count = sum(ever_death==1),
    death_probability = death_count / total_count
  )


death_prob_smoking_clean <- death_prob_smoking %>%
   ungroup() %>%
   select(age_and_local, female, death_probability) %>%
   pivot_wider(names_from = female, values_from = death_probability, names_prefix = "sex_")

colnames(death_prob_smoking_clean) <- c("Age", "Male", "Female")


common_ages <- intersect(death_prob_smoking_clean$Age, Canadalifetables_df$Age)

# filter both so only include the rows with matching Age
death_prob_smoking_filtered <- death_prob_smoking_clean[death_prob_smoking_clean$Age %in% common_ages, ]
Canadalifetables_filtered <- Canadalifetables_df[Canadalifetables_df$Age %in% common_ages, ]

combined_data_long_smoking <- bind_rows(
  death_prob_smoking_filtered %>% mutate(Source = "epicR"),
  Canadalifetables_filtered %>% mutate(Source = "Canada Life Tables")
) %>%
  pivot_longer(cols = c("Male", "Female"), names_to = "Sex", values_to = "Death_Probability") %>%
  filter(Age > 40)  

ggplot(combined_data_long_smoking, aes(x = Age, y = Death_Probability, fill = Source)) +
  geom_col(position = "dodge", width = 1) +  
  facet_wrap(~Sex) +
  labs(
    title = "Comparison of epicR Death Probability vs. Canada Life Tables (time_horizon =1)",
    x = "Age",
    y = "Death Probability",
    fill = "Source:"
  ) +
  theme_minimal()+
  theme(
    legend.position = "top",
    legend.justification = "center",
    plot.title = element_text(hjust = 0.5, margin = margin(b = 10))
  )


```




